Iterative Monomorphisation

### 2025-01-22 iterative_monomorphisation_lncs.tex
changed typo in git file name

\documentclass[]{ceurart}

@@

\SetNlSty{bfseries}{\color{black}}{}

@@

\newdefinition{definition}{Definition}
\newdefinition{example}[definition]{Example}

@@

\begin{document}

\copyrightyear{2024}
\copyrightclause{Copyright for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0)}

\conference{Submitted draft} % TODO update

@@

\author[1,2]{Tanguy Bozec}[%
email=tanguy.bozec@ens-paris-saclay.fr,
]
\author[2]{Jasmin Blanchette}[%
email=jasmin.blanchette@ifi.lmu.de,
%url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
]
\address[1]{ENS Paris-Saclay, Université Paris-Saclay, France}
\address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}

\begin{abstract}
Monomorphisation can be used to extend monomorphic provers to support polymorphic logics. We propose an iterative approach, which is necessarily incomplete but which works well in practice. It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation demonstrates that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.
\end{abstract}

\begin{keywords}
   Polymorphism\sep
   monomorphism\sep
   automatic theorem proving
\end{keywords}

@@

@@

One of the main applications of automatic theorem provers is to provide automation to users of proof assistants. Many proof assistants, such as HOL4 \cite{slind-norrish-2008}, HOL Light \cite{harrison-2009}, and
Isabelle/HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed only at the top level of formulae. On the other hand, many automatic provers operate only on monomorphic logics. One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire \cite{bhayat-reger-2020}. This, however, entails a lot of work that needs to be redone for every prover.

The alternative is to translate polymorphic problems to monomorphic problems.
One approach is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
we refer to Blanchette et al.\ \cite[Section~9]{mono-trans} for an overview of this approach.
Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme \cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating the formulae's type variables with concrete types.

By a type version of the compactness theorem, we have that in first order logic, given a polymorphic formula \(\varphi\), there exist an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, such a set cannot be computed \cite[Theorem 1]{expr-poly-types}. As a result, any monomorphisation method based on instantiation of type variables is
bound to be incomplete.

B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL. This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers. However, it is is documented only as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
A similar algorithm appears to be implemented in the \texttt{MESON} tactic \cite{harrison-1996} of HOL Light, but it is undocumented.

@@

We implemented iterative monomorphisation in Zipperposition \cite{zipp}, a higher order prover written in OCaml. Although Zipperposition is polymorphic, it uses the monomorphic prover E \cite{e} as a backend. This means that E can now be used with polymorphic problems. Moreover, our implementation of the algorithm in Zipperposition can be used as a preprocessor to interface with other stand-alone provers.

Our empirical evaluation on TPTP \cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
\begin{enumerate}
\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?

@@

\item For Leo-III \cite{leo-iii} and Vampire \cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
\end{enumerate}

@@

This algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic. In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.

@@

\begin{definition}
Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_n\mapsto\upsilon_n\}\) are \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
\end{definition}

@@

   \item For each formula \(\varphi \in \Phi\) and for each symbol \(f\) occurring in \(\varphi\):
   \begin{enumerate}
    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\) and each tuple \((\upsilon_1, \dots, \upsilon_n) \in N(\varphi)(f)\),
     for each \(i\), match \(\upsilon_i\) against \(\tau_i\), yielding the substitution \(\sigma_i\) in case of success.

@@

\begin{example}Consider the following problem:
\begin{enumerate}
   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
\end{enumerate}
%
The first iteration matches \(\alpha\) against \(\ty{int}\) for $\sym{p}$, generating the formula
%
\begin{enumerate}
   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
\end{enumerate}
%
The second iteration matches \(\alpha\) against \(\ty{list}(\ty{int})\), leading to the formula
%
\begin{enumerate}
   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
\end{enumerate}
%
Similarly the third iteration adds
%
\begin{enumerate}
   \item[5.] \(\forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,}\; \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
\end{enumerate}

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no general support for $n$-ary type constructors.
% \cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

   Depending on the shape and size of the input problem and the number of iterations performed, the doubly exponential growth may be problematic. Introducing bounds addresses this potential issue. The limit on the number of new monomorphic type argument tuples is \(\min(\max(\MonoMult\cdot m, \MonoFloor), \MonoCap)\), where \(m\) is the total number of monomorphic type argument tuples. The components of this limit are
\begin{enumerate}
   \item \MonoCap, an absolute limit on the total number of new type argument tuples;
   \item \MonoMult, which is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(m\) of monomorphic type argument tuples;
   \item \MonoFloor, which balances out \MonoMult, preventing \MonoMult from inhibiting new type argument tuple generation if \(m\) is too low.
\end{enumerate}

@@

The \emph{generate\_mono\_formulae} function is given in Figure~\ref{gen_formulae}.
It generates monomorphising substitutions and applies them to the polymorphic formula of the input problem that they instantiate. Since the substitutions are monomorphising relative to the formula they applied to, the resulting formulae are monomorphic. The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae. It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of formulae it is given.

@@

The benchmarks are taken from version 8.2.0 of the TPTP library \cite{tptp}. The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread. This corresponds to the default time limit in Sledgehammer.

@@

\begin{table}[th]
\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & &&& \multicolumn{6}{c}{cap} \\
   & \multicolumn{4}{c}{500} &\multicolumn{4}{c}{1000} & \multicolumn{4}{c}{\(\infty\)}\\
   \cmidrule(l){2-13}
   & &&& \multicolumn{6}{c}{floor} \\
   \multirow{1}{2em}{mult} & 0 & 50 & 100 & 200& 0 & 50 & 100 & 200& 0 & 50 & 100 & 200\\
    \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
    0       &\bf{178}& 161 & 161 & 156 & 178 & 160 & 160 & 156 & 178 & 161 & 160 & 156 \\
    1          & 155 & 155 & 155 & 158 & 153 & 154 & 154 & 156 & 154 & 154 & 155 & 155 \\
    2          & 154 & 154 & 153 & 154 & 153 & 153 & 154 & 152 & 154 & 153 & 154 & 154 \\
    \(\infty\) & 153 & 154 & 153 & 155 & 155 & 153 & 154 & 156 & 159 & 160 & 161 & 161 \\
    \bottomrule
\end{tabular}
\label{mono_ty_args}
\caption{Evaluation of bounds for monomorphic type argument generation}
\end{table}

@@

\begin{table}[th]
\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & &&& \multicolumn{6}{c}{cap} \\
   & \multicolumn{4}{c}{500} &\multicolumn{4}{c}{1000} & \multicolumn{4}{c}{\(\infty\)}\\
   \cmidrule(l){2-13}
   & &&& \multicolumn{6}{c}{floor} \\
   \multirow{1}{3em}{mult} & 0 & 10 & 50 & 100& 0 & 10 & 50 & 100& 0 & 10 & 50 & 100\\
    \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
    0         &125&\bf{184}& 182 & 177 & 125 & 184 & 182 & 177 & 125 & 184 & 182 & 177 \\
    0.5        & 176 & 184 & 182 & 177 & 176 & 184 & 182 & 177 & 176 & 184 & 182 & 177 \\
    1          & 182 & 181 & 178 & 177 & 182 & 181 & 178 & 177 & 182 & 181 & 178 & 177 \\
    \(\infty\) & 173 & 174 & 174 & 174 & 174 & 174 & 173 & 173 & 125 & 125 & 125 & 125 \\
    \bottomrule
\end{tabular}
\label{nmon_ty_args}
\caption{Evaluation of bounds for non-monomorphic type argument generation}
\end{table}

%The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
The substitution generation phase occurs once all type arguments have been generated. The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``separation'' heuristic is used.
% TODO find a better term than "separation heuristic" maybe separation ?
\begin{table}[th]
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{5em}}@{}}
   \toprule
   & \multicolumn{3}{c}{substitution ordering} \\
   \multirow{1}{5em}{mono subst} & age & random & separation\\
   \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
   2  & 161 & 178 & 175 \\
   5  & 161 & 178 & 180 \\
   7  & 161 & 178 & 182 \\
   10 & 161 & 178 &\bf{184} \\
   \bottomrule
\end{tabular}
\label{subst_gen_table}
\caption{Evaluation of bounds for substitution generation}
\end{table}

@@

\begin{table}[th]
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & \multicolumn{4}{c}{formula multiplier}\\
   \multirow{1}{5.2em}{formula cap} & 1 & 2 & 3 & \(\infty\)\\
   \midrule
   500       &\bf{184}& 184 & 184 & 183 \\
   2000         & 184 & 184 & 184 & 184 \\
   \(\infty\)   & 168 & 178 & 183 & 125 \\
   \bottomrule
\end{tabular}
\label{pb_size}
\caption{Evaluation of bounds directly related to the size of the output problem}
\end{table}

@@

\begin{table}[th]
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & \multicolumn{4}{c}{mono time} \\
   \multirow{1}{5em}{num.\ loops} & 5 & 10 & 20 & 30\\
   \midrule
   1     & 183 & 184 & 183 & 183  \\
   2  &\bf{186}& 186 & 186 & 185  \\
   3     & 186 & 186 & 186 & 185  \\
   4     & 186 & 186 & 186 & 185  \\
   5     & 185 & 185 & 185 & 184  \\
   \bottomrule
\end{tabular}
\label{mono_time}
\caption{Evaluation of parameters related to the depth of monomorphisation}
\end{table}

@@

\begin{table}[th]
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & \multicolumn{4}{c}{E call point} \\
   \multirow{1}{4em}{E timeout} & 0 & 0.1 & 0.2 & 0.3\\
   \midrule
   2      & 180 & 143 & 132 & 124 \\
   5   &\bf{185}& 142 & 134 & 125 \\
   10     & 184 & 143 & 132 & 125 \\
   20     & 184 & 137 & 133 & 125 \\
   30     & 182 & 133 & 134 & 125 \\
   \bottomrule
\end{tabular}
\label{e_settings}
\caption{Evaluation of parameters related to the E prover}
\end{table}

@@

\begin{table}[ht]
\centering\begin{tabular}{@{}lccc@{}}
   \toprule
   & without E & with E & Union \\
   \midrule
   500 problem benchmark   & 160 & 198 & 207 \\ % -9 +47
   1034 problem benchmark & 337 & 410 & 434 \\ % -24 +97
   \bottomrule
   \label{zipp_eval}
\end{tabular}
\caption{Evaluation of Zipperposition without E vs. with E}
\end{table}

>>>>

\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}

@@

\begin{document}
\SetNlSty{bfseries}{\color{black}}{}

@@

%\newtheorem{definition}{Definition}
%\newtheorem{example}[definition]{Example}

@@

%\copyrightyear{2024}
%\copyrightclause{Copyright for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0)}
%
%\conference{Submitted draft} % TODO update

@@

\author{Tanguy Bozec\inst{1,2} \and
Jasmin Blanchette\inst{2}}

\institute{ENS Paris-Saclay, Université Paris-Saclay, France \and
Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}

@@

\begin{abstract}
Monomorphisation can be used to extend some monomorphic provers to support polymorphic logics. We propose an iterative approach, which is necessarily incomplete but which works well in practice. It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation demonstrates that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.

\keywords{Polymorphism  \and Monomorphism    \and Automatic theorem proving}
\end{abstract}

@@

One of the main applications of automatic theorem provers is to provide automation to users of proof assistants. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and
Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed only at the top level of formulae. On the other hand, many automatic provers operate only on monomorphic logics. One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}. This, however, entails a lot of work that needs to be redone for every prover.

The alternative is to translate polymorphic problems to monomorphic problems.
One approach is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
we refer to Blanchette et al.~\cite[Section~9]{mono-trans} for an overview of this approach.
Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating the formulae's type variables with concrete types.

By a type version of the compactness theorem, we have that in first order logic, given a polymorphic formula \(\varphi\), there exist an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, such a set cannot be computed~\cite[Theorem 1]{expr-poly-types}. As a result, any monomorphisation method based on instantiation of type variables is
bound to be incomplete.

B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd} in Isabelle/HOL. This implementation is also used by Sledgehammer~\cite{judgement,hammer} to interface with superposition based automatic theorem provers. However, it is is documented only as a single subsection in his PhD thesis~\cite[Section 2.2.1]{sb-phd}.
A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.

@@

We implemented iterative monomorphisation in Zipperposition~\cite{zipp}, a higher order prover written in OCaml. Although Zipperposition is polymorphic, it uses the monomorphic prover E~\cite{e} as a backend. This means that E can now be used with polymorphic problems. Moreover, our implementation of the algorithm in Zipperposition can be used as a preprocessor to interface with other stand-alone provers.

Our empirical evaluation on TPTP~\cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
\begin{enumerate}
\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?

@@

\item For Leo-III~\cite{leo-iii} and Vampire~\cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
\end{enumerate}

@@

This algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic. In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.

@@

\begin{definition}
Given two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_n\mapsto\upsilon_n\}\) they are said to be \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
\end{definition}

@@

   \item For each formula \(\varphi \in \Phi\) and for each symbol \(f\) occurring in \(\varphi\):
   \begin{enumerate}
    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\) and \((\upsilon_1, \dots, \upsilon_n) \in N(\varphi)(f)\),
     for each \(i\), match \(\upsilon_i\) against \(\tau_i\), yielding the substitution \(\sigma_i\) in case of success.

@@

\begin{example}Consider the following problem:
\begin{enumerate}
   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
\end{enumerate}
%
The first iteration matches \(\alpha\) against \(\ty{int}\) for $\sym{p}$, generating the formula
%
\begin{enumerate}
   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
\end{enumerate}
%
The second iteration matches \(\alpha\) against \(\ty{list}(\ty{int})\), leading to the formula
%
\begin{enumerate}
   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
\end{enumerate}
%
Similarly the third iteration adds
%
\begin{enumerate}
   \item[5.] 
         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
\end{enumerate}

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no general support for $n$-ary type constructors.
%~\cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

   Depending on the shape and size of the input problem and the number of iterations performed, the doubly exponential growth may be problematic. Introducing bounds addresses this potential issue. The limit on the number of newly generated monomorphic type argument tuples is \(\min(\max(\MonoMult\cdot m, \MonoFloor), \MonoCap)\), where \(m\) is the total number of monomorphic type argument tuples. The components of this limit are
\begin{enumerate}
   \item \MonoCap, an absolute limit on the total number of new type argument tuples;
   \item \MonoMult, which is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(m\) of monomorphic type argument tuples;
   \item \MonoFloor, which balances out \MonoMult, preventing \MonoMult from inhibiting new type argument tuple generation if \(m\) is too low.
\end{enumerate}

@@

The \emph{generate\_mono\_formulae} function is given in Figure~\ref{gen_formulae}.
It generates monomorphising substitutions and applies them to the polymorphic formula of the input problem that they instantiate. Since the substitutions are monomorphising relative to the formula they applied to, the resulting formulae will be monomorphic. The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae. It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of formulae it is given.

@@

The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}. The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread. This corresponds to the default time limit in Sledgehammer.

@@

\begin{table}[th]
\caption{Evaluation of bounds for monomorphic type argument generation}
\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{2.5em}}@{}}
   \toprule
   & &&& \multicolumn{6}{c}{cap} \\
   & \multicolumn{4}{c}{500} &\multicolumn{4}{c}{1000} & \multicolumn{4}{c}{\(\infty\)}\\
   \cmidrule(l){2-13}
   & &&& \multicolumn{6}{c}{floor} \\
   \multirow{1}{2.5em}{mult} & 0 & 50 & 100 & 200& 0 & 50 & 100 & 200& 0 & 50 & 100 & 200\\
    \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
    0       &\bf{178}& 161 & 161 & 156 & 178 & 160 & 160 & 156 & 178 & 161 & 160 & 156 \\
    1          & 155 & 155 & 155 & 158 & 153 & 154 & 154 & 156 & 154 & 154 & 155 & 155 \\
    2          & 154 & 154 & 153 & 154 & 153 & 153 & 154 & 152 & 154 & 153 & 154 & 154 \\
    \(\infty\) & 153 & 154 & 153 & 155 & 155 & 153 & 154 & 156 & 159 & 160 & 161 & 161 \\
    \bottomrule
\end{tabular}
\label{mono_ty_args}
\end{table}

@@

\begin{table}[th]
\caption{Evaluation of bounds for non-monomorphic type argument generation}
\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{2.5em}}@{}}
   \toprule
   & &&& \multicolumn{6}{c}{cap} \\
   & \multicolumn{4}{c}{500} &\multicolumn{4}{c}{1000} & \multicolumn{4}{c}{\(\infty\)}\\
   \cmidrule(l){2-13}
   & &&& \multicolumn{6}{c}{floor} \\
   \multirow{1}{3em}{mult} & 0 & 10 & 50 & 100& 0 & 10 & 50 & 100& 0 & 10 & 50 & 100\\
    \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
    0         &125&\bf{184}& 182 & 177 & 125 & 184 & 182 & 177 & 125 & 184 & 182 & 177 \\
    0.5        & 176 & 184 & 182 & 177 & 176 & 184 & 182 & 177 & 176 & 184 & 182 & 177 \\
    1          & 182 & 181 & 178 & 177 & 182 & 181 & 178 & 177 & 182 & 181 & 178 & 177 \\
    \(\infty\) & 173 & 174 & 174 & 174 & 174 & 174 & 173 & 173 & 125 & 125 & 125 & 125 \\
    \bottomrule
\end{tabular}
\label{nmon_ty_args}
\end{table}

%The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
The substitution generation phase occurs once all type arguments have been generated. The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``separation'' heuristic is used.
% TODO find a better term than "separation heuristic" maybe separation ?
\begin{table}[th]
\caption{Evaluation of bounds for substitution generation}
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & \multicolumn{3}{c}{substitution ordering} \\
   \multirow{1}{6em}{mono subst} & age & random & separation\\
   \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
   2  & 161 & 178 & 175 \\
   5  & 161 & 178 & 180 \\
   7  & 161 & 178 & 182 \\
   10 & 161 & 178 &\bf{184} \\
   \bottomrule
\end{tabular}
\label{subst_gen_table}
\end{table}

@@

\begin{table}[th]
\caption{Evaluation of bounds directly related to the size of the output problem}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{3em}}@{}}
   \toprule
   & \multicolumn{4}{c}{formula multiplier}\\
   \multirow{1}{5.2em}{formula cap} & 1 & 2 & 3 & \(\infty\)\\
   \midrule
   500       &\bf{184}& 184 & 184 & 183 \\
   2000         & 184 & 184 & 184 & 184 \\
   \(\infty\)   & 168 & 178 & 183 & 125 \\
   \bottomrule
\end{tabular}
\label{pb_size}
\end{table}

@@

\begin{table}[th]
\caption{Evaluation of parameters related to the depth of monomorphisation}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{2.5em}}@{}}
   \toprule
   & \multicolumn{4}{c}{mono time} \\
   \multirow{1}{6em}{num.\ loops} & 5 & 10 & 20 & 30\\
   \midrule
   1     & 183 & 184 & 183 & 183  \\
   2  &\bf{186}& 186 & 186 & 185  \\
   3     & 186 & 186 & 186 & 185  \\
   4     & 186 & 186 & 186 & 185  \\
   5     & 185 & 185 & 185 & 184  \\
   \bottomrule
\end{tabular}
\label{mono_time}
\end{table}

@@

\begin{table}[th]
\caption{Evaluation of parameters related to the E prover}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{3em}}@{}}
   \toprule
   & \multicolumn{4}{c}{E call point} \\
   \multirow{1}{5em}{E timeout} & 0 & 0.1 & 0.2 & 0.3\\
   \midrule
   2      & 180 & 143 & 132 & 124 \\
   5   &\bf{185}& 142 & 134 & 125 \\
   10     & 184 & 143 & 132 & 125 \\
   20     & 184 & 137 & 133 & 125 \\
   30     & 182 & 133 & 134 & 125 \\
   \bottomrule
\end{tabular}
\label{e_settings}
\end{table}

@@

\begin{table}[ht]
\caption{Evaluation of Zipperposition without E vs. with E}
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & without E & with E & union \\
   \midrule
   500 problem benchmark   & 160 & 198 & 207 \\ % -9 +47
   1034 problem benchmark & 337 & 410 & 434 \\ % -24 +97
   \bottomrule
   \label{zipp_eval}
\end{tabular}
\end{table}

We compare the performance of two instances of Zipperposition. The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configuration in succession. Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic. Because all given problems are non-monomorphic, this instance is labeled ``without E''. The second instance, is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase. This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately. The results of these evaluations do not suggest that any overfitting took place during the option optimisation phase.
